Relative parabolic factorizations of Chevalley groups

### 2015-07-29 paper.tex
Small fixes in the "Chevalley group" section

In this paper we work with irreducible representations of Chevalley groups corresponding to the highest weight
$\varpi_1$ (in the case of a classical $\Phi_l$) or $\varpi_l$ (in the case $\Phi_\ell=\rE_\ell$). 
The former are the natural vector representations of classical groups acting on free modules $V=R^n$ of dimension 
$n=\ell+1, 2\ell+1, 2\ell,2\ell$ for $\Phi=\rA_\ell,\rB_\ell,\rC_\ell,\rD_\ell$.
The latter representations have dimension $n=27, 56$ for $\ell=6,7$.

%TODO: Rewrite this using better terminology.
For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$ and by $V_\lambda$ the weight subspace corresponding to $\lambda\in\Lambda(\pi)$.
For any of the representations described above (with the sole exception of $(\rE_8, \varpi_8)$) all weight subspaces $V_\lambda$ are one-dimensional. 
This allows us to choose some basis $E_\pi=\{e_\lambda\}_{\lambda\in\Lambda(\pi)}$.

It will be convenient to use standard numbering for the weights of natural representations of classical groups (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, \ell+1 & \text{in the case} & \Phi =\rA_\ell, \\
  1,2,\ldots \ell, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots \ell, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.

For a weight $\lambda\in \Lambda(\pi)$ we denote by $(-)_\lambda\colon V\rightarrow R$ the coordinate function corresponding to $\lambda$.
In other words,  $v_\lambda$ is the coefficient of the basis vector $e_\lambda$ in the expansion of $v$ in $E_\pi$, i.e. $v=\sum_{\lambda\in \Lambda(\pi)} v_\lambda e_\lambda$.
We denote by $v^+$ the basis vector corresponding to highest weight.
For classical $\Phi$ we have $v^+=e_1$.

@@

\subsection{Elementary subgroup}
Recall from~\cite{St78, VP} that for $\alpha\in \Phi$, $\xi\in R$ one can define certain elements $t_{\alpha}(\xi)$ of $\G_{sc}(\Phi, R)$ called {\it elementary root unipotents}.
These elements satisfy well-known Steinberg relations.
\begin{equation}\label{rel:add} t_\alpha(s) t_\alpha(t) = t_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF} [t_\alpha(s),  t_\beta(t)] = \left(\prod\limits t_{i\alpha + j\beta}(N_{\alpha,\beta, i, j}s^i t^j)\right),\quad \alpha\neq-\beta, N_{\alpha, \beta, i, j}\in\mathbb{Z}.\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha, \beta, i, j}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

For $\xi\in R^*$ set $w_\alpha(\xi) := x_\alpha(\xi) x_{-\alpha}(-\xi^{-1}) x_{\alpha}(\xi).$
If $\rk(\Phi)\geq 2$ the following relation is a consequence of \ref{rel:add}--\ref{rel:CCF}:
\begin{equation}\label{rel:R3} w_\alpha(\xi) t_{\beta}(s) w_\alpha(\xi)^{-1} =
t_{\sigma\ssub{\alpha}\beta} (\eta_{\alpha, \beta}\xi^{-\langle\beta,\alpha \rangle} s),\ \xi,s\in R. \end{equation}
where coefficients $\eta_{\alpha, \beta} = \pm 1$ (see~\cite[\S13]{VP} for more details on $\eta_{\alpha, \beta}$).

\begin{rem} Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $t_{\lambda_1,\lambda_2}(\xi)$ instead of $t_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ we have $t_{1,2}(\xi)=t_{1-2}(\xi)=t_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = t_{\alpha_1}(\xi)$. \end{rem}

Let $I\leq R$ be an ideal. Denote by $\E(\Phi, I)$ the subgroup of $\G_{sc}(\Phi, R)$ generated by elementary root unipotents $t_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in I$.
In the special case $I=R$ the subgroup $\E(\Phi, R)$ is called the {\it elementary subgroup} of the Chevalley group $\G_{sc}(\Phi, R)$.
Taddei's theorem asserts that $\E(\Phi, R)$ is a normal subgroup of $\G_{sc}(\Phi, R)$ provided $\Phi$ is an irreducible root system of rank $\geq 2$.

We denote by $\E(\Phi, R, I)$ the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, in other words $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.
Denote by $z_\alpha(s, \xi)$ the element $t_\alpha(s)^{t\ssub{-\alpha}(\xi)}$.
The following result of L.~Vaserstein and J.~Tits describes a set of generators of $\E(\Phi, R, I)$ (see e.\,g.~\cite[Theorem~2]{Va86}).

\begin{thm} Let $\Phi$ be an irreducible root system of rank $\geq 2$, then the relative elementary subgroup $\E(\Phi, R, I)$
is generated by elements $z_\alpha(s, \xi)$, $\alpha\in \Phi$, $s\in I$, $\xi\in R$.
\end{thm}

In fact, one can choose a more economical set of generators for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the following two families of elements:
\begin{itemize}
 \item $t_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$. \end{itemize} \end{thm}

For a special subset of roots $S\subseteq \Phi$ we denote by $\U(S, I)$ the subgroup spanned by all $t_{\alpha}(s)$ for $s\in I$, $\alpha\in S$.
The subgroup $\U(\Sigma_J, I)$ is normalized by $\E(\Delta_J, R)$, hence the Minkowski product set $\EP_J(R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I)$ is a subgroup. 
The following two equalities will be referred to in the sequel as {\it Levi decomposition}: 
$$\EP_J(R, I) = \U(\Sigma_J, I) \cdot \E(\Delta_J, R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I).$$

>>>>

In this paper we work with the irreducible representations of Chevalley groups corresponding to the highest weight
$\varpi_1$ (in the case of a classical $\Phi_\ell$) or $\varpi_\ell$ (in the case $\Phi_\ell=\rE_\ell$). 
The former are the natural vector representations of the classical groups acting on the free modules $V=R^n$ of dimension 
$n=\ell+1, 2\ell+1, 2\ell,2\ell$ for $\Phi=\rA_\ell,\rB_\ell,\rC_\ell,\rD_\ell$ respectively.
The latter are the microweight representations of dimension $n=27, 56$ for $\ell=6,7$.

%TODO: Rewrite this using better terminology.
For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda=\Lambda(\pi)$ the set of weights of $\pi$ and by $V_\lambda$ the weight subspace corresponding to $\lambda\in\Lambda$.
For any of the representations described above (with the sole exception of $(\rE_8, \varpi_8)$) all weight subspaces $V_\lambda$ are one-dimensional. 
This allows us to choose a basis $E_\pi=\{e_\lambda\}_{\lambda\in\Lambda}$.

It will be convenient to use the standard numbering for the weights of the natural representations of the classical groups (cp.~\cite[§1B]{St78}):
\[\begin{array}{cll}
  1,2,\ldots, \ell+1 & \text{in the case} & \Phi =\rA_\ell, \\
  1,2,\ldots \ell, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots \ell, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}\]
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.

For a weight $\lambda\in \Lambda$ we denote by $(-)_\lambda\colon V\to R$ the coordinate function corresponding to $\lambda$.
In other words,  $v_\lambda$ is the coefficient of the basis vector $e_\lambda$ in the expansion of $v$ in $E_\pi$, i.e. $v=\sum_{\lambda\in \Lambda} v_\lambda e_\lambda$.
We denote by $v^+$ the basis vector corresponding to the highest weight.
For classical $\Phi$ we have $v^+=e_1$.

@@

\subsection{Elementary subgroup}
Recall from~\cite{St78, VP} that for $\alpha\in \Phi$, $\xi\in R$ one can define certain elements $x_{\alpha}(\xi)$ of $\G_{sc}(\Phi, R)$ called {\it elementary root unipotents}.
These elements satisfy the well-known Steinberg relations:
\begin{equation}\label{rel:add}
x_\alpha(s) x_\alpha(t) = x_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF}
\left[x_\alpha(s),  x_\beta(t)\right] = \prod x_{i\alpha + j\beta}\left(N_{\alpha\beta ij}\, s^i t^j\right),\quad \alpha\neq-\beta, \quad N_{\alpha\beta ij}\in\mathbb{Z}.
\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha\beta ij}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

For $\xi\in R^*$ set $w_\alpha(\xi) = x_\alpha(\xi) x_{-\alpha}(-\xi^{-1}) x_{\alpha}(\xi).$
If $\rk(\Phi)\geq 2$ the following relation is a consequence of \ref{rel:add}--\ref{rel:CCF}:
\begin{equation}\label{rel:R3}
w_\alpha(\xi) x_{\beta}(s) w_\alpha(\xi)^{-1} =
x_{\sigma\ssub{\alpha}\beta} \left(\eta_{\alpha, \beta}\xi^{-\langle\beta,\alpha \rangle} s\right),\quad \xi,s\in R.
\end{equation}
where coefficients $\eta_{\alpha, \beta} = \pm 1$ (see~\cite[\S13]{VP} for more details on $\eta_{\alpha, \beta}$).

\begin{rem} Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $x_{\lambda_1,\lambda_2}(\xi)$ instead of $x_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ we have $x_{1,2}(\xi)=x_{1-2}(\xi)=x_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = x_{\alpha_1}(\xi)$. \end{rem}

Let $I\leq R$ be an ideal. Denote by $\E(\Phi, I)$ the subgroup of $\G_{sc}(\Phi, R)$ generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in I$.
In the special case $I=R$ the subgroup $\E(\Phi, R)$ is called the {\it elementary subgroup} of the Chevalley group $\G_{sc}(\Phi, R)$.
Taddei's theorem asserts that $\E(\Phi, R)$ is a normal subgroup of $\G_{sc}(\Phi, R)$ provided $\Phi$ is an irreducible root system of rank $\geqslant 2$.

We denote by $\E(\Phi, R, I)$ the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, in other words $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.
Denote by $z_\alpha(s, \xi)$ the element $x_\alpha(s)^{x\ssub{-\alpha}(\xi)}$.
The following result of L.~Vaserstein and J.~Tits describes a set of generators of $\E(\Phi, R, I)$ (see e.\,g.~\cite[Theorem~2]{Va86}).

\begin{thm} Let $\Phi$ be an irreducible root system of rank $\geq 2$, then the relative elementary subgroup $\E(\Phi, R, I)$
is generated by the elements $z_\alpha(s, \xi)$, $\alpha\in \Phi$, $s\in I$, $\xi\in R$.
\end{thm}

In fact, one can choose a more economical set of generators for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with the special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the union of the following two families of elements:
\begin{itemize}
 \item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$. \end{itemize} \end{thm}

For a special subset of roots $S\subseteq \Phi$ we denote by $\U(S, I)$ the subgroup spanned by all $x_{\alpha}(s)$ for $s\in I$, $\alpha\in S$.
The subgroup $\U(\Sigma_J, I)$ is normalized by $\E(\Delta_J, R)$, hence the Minkowski product set $\EP_J(R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I)$ is a subgroup. 
The following two equalities will be referred to in the sequel as {\it Levi decomposition}: 
$$\EP_J(R, I) = \U(\Sigma_J, I) \cdot \E(\Delta_J, R, I) = \E(\Delta_J, R, I) \cdot \U(\Sigma_J, I).$$
